A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2024-03-24 paper.tex
R2 submitted

\articletype{ARTICLE TEMPLATE}

@@

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the availability of many numerical hypothesis test
procedures designed to use residuals to assess problems with a model
fit. Here we provide evidence for why this is good advice using data
from a visual inference experiment. We show how conventional tests are
too sensitive, which means that too often the conclusion would be that
the model fit is inadequate. The experiment uses the lineup protocol
which puts a residual plot in the context of null plots. This helps
generate reliable and consistent reading of residual plots for better
model diagnosis. It can also help in an obverse situation where a
conventional test would fail to detect a problem with a model due to
contaminated data. The lineup protocol also detects a range of
departures from good residuals simultaneously.
\end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting the data plot (at position $2^2 + 2$, exhibiting non-linearity).}\label{fig:first-example-lineup}
\end{figure}

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{departures-from-good-residual-plots}{%
\subsection{Departures from good residual
plots}\label{departures-from-good-residual-plots}}

@@

\caption{Example residual vs fitted value plots (red line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot.}\label{fig:residual-plot-common-departures}
\end{figure}

@@

\hypertarget{conventionally-testing-for-departures}{%
\subsection{Conventionally testing for
departures}\label{conventionally-testing-for-departures}}

@@

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, BP and SW tests applied to the residual plots in Figure
\ref{fig:residual-plot-common-departures}. The RESET test and BP test
were computed using the \texttt{resettest} and \texttt{bptest} functions
from the R package \texttt{lmtest}, respectively. The SW test was
computed using the \texttt{shapiro.test} from the core R package
\texttt{stats}. \footnote{Although we didn't use it, it's useful to know
  that the R package \texttt{skedastic} \citep{skedastic} also contains
  a large collection of functions to test for heteroskedasticity.} The
RESET test requires the selection of a power parameter.
\citet{ramsey1969tests} recommends a power of four, which we adopted in
our analysis.

For residual plots in Figure \ref{fig:residual-plot-common-departures},
we would expect the RESET test for non-linearity to reject residual plot
B, the BP test for heteroskedasticity to reject the residual plot C, and
the SW test for non-normality to reject residual plot D, which they all
do and all tests also correctly fail to reject residual plot A.
Interestingly, the BP and SW tests also reject the residual plots
exhibiting structure that they weren't designed for.
\citet{cook1982residuals} explain that most residual-based tests for a
particular departure from the model assumptions are also sensitive to
other types of departures. This could be considered a Type III error
\citep{kimball1957errors}, where the null hypothesis of good residuals
is correctly rejected but for the wrong reason. Also, some types of
departure can have elements of other types of departure, for example,
non-linearity can appear like heteroskedasticity. Additionally, other
data problems such as outliers can trigger rejection (or not) of the
null hypothesis \citep{cook1999applied}.

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (A good discussion can be found in
\citet{kirk1996practical}.) While such rejections may be statistically
correct, their sensitivity may render the results impractical. A key
goal of residual plot diagnostics is to identify potential issues that
could lead to incorrect conclusions or errors in subsequent analyses,
but minor defects in the model are unlikely to have a significant impact
and may be best disregarded for practical purposes. The experiment
discussed in this paper specifically addresses this tension between
statistical and practical significance.

@@

\hypertarget{visual-test-procedure-based-on-lineups}{%
\subsection{Visual test procedure based on
lineups}\label{visual-test-procedure-based-on-lineups}}

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(2^2 + 2\) is identifiable, then
it is evidence for the rejection of \(H_0\). In fact, the actual
residual plot is obtained from a misspecified regression model with
missing non-linear terms.

@@

\hypertarget{significance-calculation}{%
\section{Calculation of statistical significance and test
power}\label{significance-calculation}}

\hypertarget{what-is-being-tested}{%
\subsection{What is being tested?}\label{what-is-being-tested}}

@@

\hypertarget{statistical-significance}{%
\subsection{\texorpdfstring{Statistical
significance\label{sig}}{Statistical significance}}\label{statistical-significance}}

@@

\hypertarget{power-of-the-tests}{%
\subsection{Power of the tests}\label{power-of-the-tests}}

@@

\hypertarget{experimental-design}{%
\section{Experimental design}\label{experimental-design}}

@@

\caption{\label{tab:model-factor-table}Levels of the factors used in data collection periods I, II, and III.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{rr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rc}
\toprule
\multicolumn{2}{c}{Non-linearity} & \multicolumn{2}{c}{Heteroskedasticity} & \multicolumn{2}{c}{Common} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
\multicolumn{1}{c}{Poly Order ($j$)} & \multicolumn{1}{c}{SD ($\sigma$)} & \multicolumn{1}{c}{Shape ($a$)} & \multicolumn{1}{c}{Ratio ($b$)} & \multicolumn{1}{c}{Size ($n$)} & \multicolumn{1}{c}{Distribution of the predictor} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
2 & 0.25 & -1 & 0.25 & 50 & Uniform\\
3 & 1.00 & 0 & 1.00 & 100 & Normal\\
6 & 2.00 & 1 & 4.00 & 300 & Skewed\\
18 & 4.00 &  & 16.00 &  & Discrete\\
 &  &  & 64.00 &  & \\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{simulating-departures-from-good-residuals}{%
\subsection{Simulating departures from good
residuals}\label{simulating-departures-from-good-residuals}}

\hypertarget{non-linearity-and-heteroskedasticity}{%
\subsubsection{Non-linearity and
Heteroskedasticity}\label{non-linearity-and-heteroskedasticity}}

@@

Data collection period II was similar to period I but focuses on
heteroskedasticity departures. We generated the heteroskedasticity
departures by setting the variance-covariance matrix of the error term
as a function of the predictor, but fitted the data with the simple
linear regression model, intentionally violated the constant variance
assumption. Visual patterns of heteroskedasticity are simulated using
three different shapes (\(a\) = -1, 0, 1) including ``left-triangle'',
``butterfly'' and ``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows
the butterfly shape as the ratio parameter (\(b\)) is changed. More
details about the simulation process are provided in the Appendix.

@@

\hypertarget{factors-common-to-both-data-collection-periods}{%
\subsubsection{Factors common to both data collection
periods}\label{factors-common-to-both-data-collection-periods}}

@@

Figure \ref{fig:different-dist} shows the non-linear pattern, a ``U''
shape, with the different fitted value distributions. We would expect
that structure in residual plots would be easier to perceive when the
predictors are uniformly distributed.

@@

\hypertarget{effect-size}{%
\subsection{Effect size}\label{effect-size}}

@@

For the purposes of our calculations we have chosen to use an approach
based on Kullback-Leibler divergence \citep{kullback1951information},
coupled with simulation. This formulation defines effect size to be

@@

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix, \(\boldsymbol{X}\) is the design
matrix, \(\boldsymbol{V}\) is the actual covariance matrix of the error
term,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contains
any higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}_n\) is the assumed covariance
matrix of the error term when \(H_0\) is true.

To compute the effect size for each lineup we simulate a sufficiently
large number of samples from the same model with the number of
observations \(n\) fixed for each sample. We then compute the effect
size for each sample and take the average as the final value. This
ensures lineups constructed with the same experimental factors will
share the same effect size.

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{power-comparison-of-the-tests}{%
\subsection{\texorpdfstring{Power comparison of the
tests\label{power-analysis}}{Power comparison of the tests}}\label{power-comparison-of-the-tests}}

Figure \ref{fig:nonlinearheterpower} present the power curves of various
tests plotted against the effect size in the residuals for non-linearity
and heteroskedasticity. In each case the power of visual test is
calculated for multiple bootstrap samples leading to the many (orange)
curves. The effect size was computed at a 5\% significance level and
plotted on a natural logarithmic scale. To facilitate visual calibration
of effect size values with the corresponding diagnostic plots, a
sequence of example residual plots with increasing effect sizes is
provided at the bottom of these figures. These plots serve as a visual
aid to help readers understand how different effect size values
translate to changes in the diagnostic plots. The horizontal lines of
dots at 0 and 1 represent the non-rejection or rejection decisions made
by visual tests for each lineup.

@@

Overall, the results show that the conventional tests are more sensitive
than the visual test. The conventional tests do have higher power for
the patterns they are designed to detect, but they typically fail to
detect other patterns unless those patterns are particularly strong. The
visual test doesn't require specifying the pattern ahead of time,
relying purely on whether the observed residual plot is detectably
different from ``good'' residual plots. They will perform equally well
regardless of the type of model defect. This aligns with the advice of
experts on residual analysis, who consider residual plot analysis to be
an indispensable tool for diagnosing model problems. What we gain from
using a visual test for this purpose is the removal of any subjective
arguments about whether a pattern is visible or not. The lineup protocol
provides the calibration for detecting patterns: if the pattern in the
data plot cannot be distinguished from the patterns in good residual
plots, then no discernible problem with the model exists.

\hypertarget{comparison-of-test-decisions-based-on-p-values}{%
\subsection{\texorpdfstring{Comparison of test decisions based on
\(p\)-values\label{p-value}}{Comparison of test decisions based on p-values}}\label{comparison-of-test-decisions-based-on-p-values}}

The power comparison demonstrates that the appropriate conventional
tests will reject more aggressively than visual tests, but we don't know
how the decisions for each lineup would agree or disagree. Here we
compare the reject or fail to reject decisions of these tests, across
all the lineups. Figure \ref{fig:p-value-comparison} shows the agreement
of the conventional and visual tests using a mosaic plot for both
non-linearity patterns and heteroskedasticity patterns. For both
patterns the lineups resulting in a rejection by the visual test are
\emph{all} also rejected by the conventional test, except for one from
the heteroskedasticity model. This reflects exactly the story from the
previous section, that the conventional tests reject more aggressively
than the visual test.

@@

Because the power curve of the visual tests are a shift to the right of
the conventional test (Figure \ref{fig:nonlinearheterpower}) we examined
whether adjusting the significance level (to .001, .0001, .00001,
\ldots) of the conventional test would generate similar decisions to
that of the visual test. Interestingly, it doesn't: despite resulting in
less rejections, neither the RESET or BP tests come to complete
agreement with the visual test (see Appendix).

\hypertarget{effect-of-amount-of-non-linearity}{%
\subsection{\texorpdfstring{Effect of amount of
non-linearity\label{nonlin-analysis}}{Effect of amount of non-linearity}}\label{effect-of-amount-of-non-linearity}}

The order of the polynomial is a primary factor contributing to the
pattern produced by the non-linearity model. Figure
\ref{fig:poly-power-uniform-j} explores the relationship between
polynomial order and power of the tests. The conventional tests have
higher power for lower orders of Hermite polynomials, and the power
drops substantially for the ``triple-U'' shape. To understand why this
is, we return to the application of the RESET test, which requires a
parameter indicating degree of fitted values to test for, and the
recommendation is to generically use four \citep{ramsey1969tests}.
However, the ``triple-U'' shape is constructed from the Hermite
polynomials using power up to 18. If the RESET test had been applied
using a higher power of no less than six, the power curve of
``triple-U'' shape will be closer to other power curves. This
illustrates the sensitivity of the conventional test to the parameter
choice, and highlights a limitation: it helps to know the data
generating process to set the parameters for the test, which is
unrealistic in practice. However, we examined this in more detail (see
Appendix) and found that there is no harm in setting the parameter
higher than four on the tests' operation for lower order polynomial
shapes. Using a parameter value of six, instead of four, yields higher
power regardless of generating process, and is recommended.

@@

\hypertarget{effect-of-shape-of-heteroskedasticity}{%
\subsection{\texorpdfstring{Effect of shape of
heteroskedasticity\label{hetero-analysis}}{Effect of shape of heteroskedasticity}}\label{effect-of-shape-of-heteroskedasticity}}

@@

\hypertarget{effect-of-fitted-value-distributions}{%
\subsection{Effect of fitted value
distributions}\label{effect-of-fitted-value-distributions}}

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests (columns) for non-linearity and heteroskedasticity patterns (rows). The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and lognormal distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

@@

Perhaps surprisingly, the visual tests have more consistent power across
the different fitted value distributions: for the non-linear pattern,
there is almost no power difference, and for the heteroskedastic
pattern, uniform and discrete have higher power than normal and
lognormal. The likely reason is that these latter two have fewer
observations in the tails where the heteroskedastic pattern needs to be
detected.

@@

\hypertarget{limitations-and-practicality}{%
\section{Limitations and
practicality}\label{limitations-and-practicality}}

@@

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

@@

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\hypertarget{supplementary-material}{%
\section*{Supplementary material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary material}

The supplementary material is available at
\url{https://github.com/TengMCing/lineup_residual_diagnostics/blob/master/appendix.pdf}.
It includes more details about the experimental setup, the derivation of
the effect size, the effect of data collection period, and the estimate
of \(\alpha\). The R package \texttt{visage} is currently available at
\url{https://github.com/TengMCing/visage}.

>>>>

\articletype{}

@@

\begin{abstract}
Regression experts consistently recommend plotting residuals for model
diagnosis, despite the availability of many numerical hypothesis test
procedures designed to use residuals to assess problems with a model
fit. Here we provide evidence for why this is good advice using data
from a visual inference experiment. We show how conventional tests are
too sensitive, which means that too often the conclusion would be that
the model fit is inadequate. The experiment uses the lineup protocol
which puts a residual plot in the context of null plots. This helps
generate reliable and consistent reading of residual plots for better
model diagnosis. It can also help in an obverse situation where a
conventional test would fail to detect a problem with a model due to
contaminated data. The lineup protocol also detects a range of
departures from good residuals simultaneously. Supplemental materials
for the article are available online.
\end{abstract}

@@

\section{Introduction}\label{introduction}

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting the data plot (at position $6$, exhibiting non-linearity).}\label{fig:first-example-lineup}
\end{figure}

\section{Background}\label{background}

\subsection{Departures from good residual
plots}\label{departures-from-good-residual-plots}

@@

\caption{Example residual vs fitted value plots (horizontal line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot.}\label{fig:residual-plot-common-departures}
\end{figure}

@@

\subsection{Conventionally testing for
departures}\label{conventionally-testing-for-departures}

@@

Table \ref{tab:example-residual-plot-table} displays the \(p\)-values
from the RESET, BP and SW tests applied to the residual plots in Figure
\ref{fig:residual-plot-common-departures}. The RESET test and BP test
were computed using the \texttt{resettest} and \texttt{bptest} functions
from the R package \texttt{lmtest}, respectively. The SW test was
computed using the \texttt{shapiro.test} from the core R package
\texttt{stats}. \footnote{Although we did not use it, it is useful to
  know that the R package \texttt{skedastic} \citep{skedastic} also
  contains a large collection of functions to test for
  heteroskedasticity.} The RESET test requires the selection of a power
parameter. \citet{ramsey1969tests} recommends a power of four, which we
adopted in our analysis.

For residual plots in Figure \ref{fig:residual-plot-common-departures},
we would expect the RESET test for non-linearity to reject residual plot
B, the BP test for heteroskedasticity to reject the residual plot C, and
the SW test for non-normality to reject residual plot D, which they all
do and all tests also correctly fail to reject residual plot A.
Interestingly, the BP and SW tests also reject the residual plots
exhibiting structure that they were not designed for.
\citet{cook1982residuals} explain that most residual-based tests for a
particular departure from the model assumptions are also sensitive to
other types of departures. This could be considered a Type III error
\citep{kimball1957errors}, where the null hypothesis of good residuals
is correctly rejected but for the wrong reason. Also, some types of
departure can have elements of other types of departure, for example,
non-linearity can appear like heteroskedasticity. Additionally, other
data problems such as outliers can trigger rejection (or not) of the
null hypothesis \citep{cook1999applied}.

With large sample sizes, hypothesis tests may reject the null hypothesis
when there is only a small effect. (A good discussion can be found in
\citet{kirk1996practical}.) While such rejections may be statistically
correct, their sensitivity may render the results impractical. A key
goal of residual plot diagnostics is to identify potential issues that
could lead to incorrect conclusions or errors in subsequent analyses,
but minor defects in the model are unlikely to have a significant impact
and may be best disregarded for practical purposes. The experiment
discussed in this paper specifically addresses this tension between
statistical significance and practical significance.

@@

\subsection{Visual test procedure based on
lineups}\label{visual-test-procedure-based-on-lineups}

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(6\) is identifiable, then it is
evidence for the rejection of \(H_0\). In fact, the actual residual plot
is obtained from a misspecified regression model with missing non-linear
terms.

@@

\section{Calculation of statistical significance and test
power}\label{significance-calculation}

\subsection{What is being tested?}\label{what-is-being-tested}

@@

\subsection{\texorpdfstring{Statistical
significance\label{sig}}{Statistical significance}}\label{statistical-significance}

@@

\subsection{Power of the tests}\label{power-of-the-tests}

@@

\section{Experimental design}\label{experimental-design}

@@

\caption{\label{tab:model-factor-table}Levels of the factors used in data collection periods I, II, and III.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{rr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rc}
\toprule
\multicolumn{2}{c}{Non-linearity} & \multicolumn{2}{c}{Heteroskedasticity} & \multicolumn{2}{c}{Common} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
\multicolumn{1}{c}{Poly Order ($j$)} & \multicolumn{1}{c}{SD ($\sigma$)} & \multicolumn{1}{c}{Shape ($a$)} & \multicolumn{1}{c}{Ratio ($b$)} & \multicolumn{1}{c}{Size ($n$)} & \multicolumn{1}{c}{Distribution of the fitted values} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
2 & 0.25 & -1 & 0.25 & 50 & Uniform\\
3 & 1.00 & 0 & 1.00 & 100 & Normal\\
6 & 2.00 & 1 & 4.00 & 300 & Skewed\\
18 & 4.00 &  & 16.00 &  & Discrete\\
 &  &  & 64.00 &  & \\
\bottomrule
\end{tabular}}
\end{table}

\subsection{Simulating departures from good
residuals}\label{simulating-departures-from-good-residuals}

\subsubsection{Non-linearity and
Heteroskedasticity}\label{non-linearity-and-heteroskedasticity}

@@

Data collection period II was similar to period I but focuses on
heteroskedasticity departures. We generated the heteroskedasticity
departures by setting the variance-covariance matrix of the error term
as a function of the predictor, but fitted the data with the simple
linear regression model, intentionally violated the constant variance
assumption. Visual patterns of heteroskedasticity are simulated using
three different shapes (\(a\) = -1, 0, 1) including ``left-triangle'',
``butterfly'' and ``right-triangle'' shapes as displayed in Figure
\ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows
the butterfly shape as the ratio parameter (\(b\)) is changed. More
details about the simulation process are provided in Appendix B.

@@

\subsubsection{Factors common to both data collection
periods}\label{factors-common-to-both-data-collection-periods}

@@

Figure \ref{fig:different-dist} shows the non-linear pattern, a ``U''
shape, with the different fitted value distributions. We would expect
that structure in residual plots would be easier to perceive when the
fitted values are uniformly distributed.

@@

\subsection{Effect size}\label{effect-size}

@@

For the purposes of our calculations we have chosen to use an approach
based on Kullback-Leibler divergence \citep{kullback1951information}.
This formulation defines effect size to be

@@

\noindent where \(\text{diag}(.)\) is the diagonal matrix constructed
from the diagonal elements of a matrix, \(\boldsymbol{X}\) is the design
matrix, \(\boldsymbol{V}\) is the actual covariance matrix of the error
term,
\(\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator,
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals where \(\boldsymbol{Z}\) contains
any higher order terms of \(\boldsymbol{X}\) left out of the regression
equation, \(\boldsymbol{\beta}_z\) contains the corresponding
coefficients, and \(\sigma^2\boldsymbol{I}_n\) is the assumed covariance
matrix of the error term when \(H_0\) is true. More details about the
effect size derivation are provided in Appendix A.

\section{Results}\label{results}

@@

\subsection{\texorpdfstring{Power comparison of the
tests\label{power-analysis}}{Power comparison of the tests}}\label{power-comparison-of-the-tests}

Figure \ref{fig:nonlinearheterpower} present the power curves of various
tests plotted against the effect size in the residuals for non-linearity
and heteroskedasticity. In each case the power of visual test is
calculated for multiple bootstrap samples leading to the many (solid
orange) curves. The effect size was computed at a 5\% significance level
and plotted on a natural logarithmic scale. To facilitate visual
calibration of effect size values with the corresponding diagnostic
plots, a sequence of example residual plots with increasing effect sizes
is provided at the bottom of these figures. These plots serve as a
visual aid to help readers understand how different effect size values
translate to changes in the diagnostic plots. The horizontal lines of
dots at 0 and 1 represent the non-rejection or rejection decisions made
by visual tests for each lineup.

@@

Overall, the results show that the conventional tests are more sensitive
than the visual test. The conventional tests do have higher power for
the patterns they are designed to detect, but they typically fail to
detect other patterns unless those patterns are particularly strong. The
visual test does not require specifying the pattern ahead of time,
relying purely on whether the observed residual plot is detectably
different from ``good'' residual plots. They will perform equally well
regardless of the type of model defect. This aligns with the advice of
experts on residual analysis, who consider residual plot analysis to be
an indispensable tool for diagnosing model problems. What we gain from
using a visual test for this purpose is the removal of any subjective
arguments about whether a pattern is visible or not. The lineup protocol
provides the calibration for detecting patterns: if the pattern in the
data plot cannot be distinguished from the patterns in good residual
plots, then no discernible problem with the model exists.

\subsection{\texorpdfstring{Comparison of test decisions based on
\(p\)-values\label{p-value}}{Comparison of test decisions based on p-values}}\label{comparison-of-test-decisions-based-on-p-values}

The power comparison demonstrates that the appropriate conventional
tests will reject more aggressively than visual tests, but we do not
know how the decisions for each lineup would agree or disagree. Here we
compare the reject or fail to reject decisions of these tests, across
all the lineups. Figure \ref{fig:p-value-comparison} shows the agreement
of the conventional and visual tests using a mosaic plot for both
non-linearity patterns and heteroskedasticity patterns. For both
patterns the lineups resulting in a rejection by the visual test are
\emph{all} also rejected by the conventional test, except for one from
the heteroskedasticity model. This reflects exactly the story from the
previous section, that the conventional tests reject more aggressively
than the visual test.

@@

Because the power curve of the visual tests are a shift to the right of
the conventional test (Figure \ref{fig:nonlinearheterpower}) we examined
whether adjusting the significance level (to .001, .0001, .00001,
\ldots) of the conventional test would generate similar decisions to
that of the visual test. Interestingly, it does not: despite resulting
in less rejections, neither the RESET or BP tests come to complete
agreement with the visual test (see Appendix A).

\subsection{\texorpdfstring{Effect of amount of
non-linearity\label{nonlin-analysis}}{Effect of amount of non-linearity}}\label{effect-of-amount-of-non-linearity}

The order of the polynomial is a primary factor contributing to the
pattern produced by the non-linearity model. Figure
\ref{fig:poly-power-uniform-j} explores the relationship between
polynomial order and power of the tests. The conventional tests have
higher power for lower orders of Hermite polynomials, and the power
drops substantially for the ``triple-U'' shape. To understand why this
is, we return to the application of the RESET test, which requires a
parameter indicating degree of fitted values to test for, and the
recommendation is to generically use four \citep{ramsey1969tests}.
However, the ``triple-U'' shape is constructed from the Hermite
polynomials using power up to 18. If the RESET test had been applied
using a higher power of no less than six, the power curve of
``triple-U'' shape will be closer to other power curves. This
illustrates the sensitivity of the conventional test to the parameter
choice, and highlights a limitation: it helps to know the data
generating process to set the parameters for the test, which is
unrealistic in practice. However, we examined this in more detail (see
Appendix A) and found that there is no harm in setting the parameter
higher than four on the tests' operation for lower order polynomial
shapes. Using a parameter value of six, instead of four, yields higher
power regardless of generating process, and is recommended.

@@

\subsection{\texorpdfstring{Effect of shape of
heteroskedasticity\label{hetero-analysis}}{Effect of shape of heteroskedasticity}}\label{effect-of-shape-of-heteroskedasticity}

@@

\subsection{Effect of fitted value
distributions}\label{effect-of-fitted-value-distributions}

@@

\caption{Comparison of power on lineups with different fitted value distributions for conventional and visual tests (columns) for non-linearity and heteroskedasticity patterns (rows). The power curves of conventional tests for non-linearity and heteroskedasticity patterns are produced by RESET tests and BP tests, respectively. Power curves of visual tests are estimated using five evaluations on each lineup. For lineups with a uniform fitted value distribution, the five evaluations are repeatedly sampled from the total eleven evaluations to give multiple power curves (solid grey). Surprisingly, the fitted value distribution has produces more variability in the power of conventional tests than visual tests. Uneven distributions, normal and skewed distributions, tend to yield lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}

@@

Perhaps surprisingly, the visual tests have more consistent power across
the different fitted value distributions: for the non-linear pattern,
there is almost no power difference, and for the heteroskedastic
pattern, uniform and discrete have higher power than normal and skewed.
The likely reason is that these latter two have fewer observations in
the tails where the heteroskedastic pattern needs to be detected.

@@

\section{Limitations and
practicality}\label{limitations-and-practicality}

@@

\section{Conclusions}\label{conclusions}

@@

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

@@

\section*{Supplementary materials}\label{supplementary-materials}
\addcontentsline{toc}{section}{Supplementary materials}

\begin{description}
\item{Appendix:} The appendix includes more details about the experiment setup, the derivation of the effect size, the effect of data collection period, and the estimate of $\alpha$. (appendix.pdf, Portable Document Format file)
\item{R package ``visage":} The R package ``visage" containing code to simulate lineups used in the experiment. The package also contains de-identified data collected from the experiment. (visage\_0.1.2.tar.gz, GNU zipped tar file)
\end{description}
